Certifying, incremental type checking

### 2012-06-13 doc/papers/LFMTP2012/paper.tex
paper

\documentclass{llncs}

\usepackage{ntheorem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{macros}
\usepackage{mathpartir}
\usepackage{url}

\allowdisplaybreaks[1]

@@

  A lightweight way to design a trusted type-checker is to let it
  return a \emph{certificate} of well-typing (for instance
  \sysname{Agda} and \sysname{GHC} adopt this architecture
  internally).  Most of the time, these type-checkers are confronted
  sequentially with very similar pieces of input: the same program,
  each time slightly modified. Yet they rebuild the entire
  certificates, even for unchanged parts of the code. Instead of
  discarding them between runs, we can take advantage of these
  certificates to improve the reactivity of type-checking, by feeding
  parts of the certificates for unchanged parts of the original
  program back into the new input, thus building an incremental
  type-checking.

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for mere ``type checking'', or explicit
``church-style'' calculi. Take for example dependent type systems,
relying on a certain notion of definitional equality on types, or type
systems with subtype polymorphism: both rely on a rule of the form
$$
\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer has to convert the declarative system to an algorithm, and
convince himself that both are equivalent. This task is highly
non-trivial and non-modular: for instance, how to infer the type of a
System \sysname{T} recursor $\recb M N x y P$, in presence of
subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
a typing derivation out of program $M$; it doesn't have to be
complete, or even sound, and \emph{(ii)} a trusted and fast kernel
$\fcheck{\md}$ to verify these derivations \wrt\ a given set of inference
rules $\Sigma$. This is a \emph{certifying} scheme (see
\cite{leroy2006compcert} for a discussion on approaches to reliable
compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$, but the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
(return an incorrect derivation). Explicitly generating the derivation
in the declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

Several generic methods have been devised for turning a batch program
into an incremental one
\cite{pugh1989incremental,acar2003selective}. Unfortunately, they all
suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates this derivation, it inserts a coercion $\cst{odd}
\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here. $\md_2$ was deriving $\vdash\s\z:\cst{nat}$ (with a
coercion since it was used as an argument to $\var{add}$); the system
thus infer the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one into the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} x y {\s{\var x}}}
  \\&\quad
  {(\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}})}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

% lambda-lifting

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solution to this question. It
is a dependently typed lambda-calculus for \emph{representing} terms,
derivations and statements of $\Pi_0^1$ meta-theorems (but not their
proofs) in a given inference system supporting hypothetical judgments
\cite{pfenning2001logical}. For these, it provides built-in support
for encoding languages with binders and hypothesis (a feature called
$\lambda$-tree syntax or HOAS \cite{pfenning1988higher}). The user
gives a \emph{signature} defining abstract syntax and inference rules
of his \emph{object language}, and gets an algorithm to verify terms
and derivations expressed in the \LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}
This framework has been used to issue certificates for a number of
practical applications like proof-carrying
code~\cite{necula1997proof}, authentication
framework~\cite{appel1999proof} \ldots %TODO ATP

@@

\begin{figure}\centering
  \begin{align*}
    K &\gequal \prd x A K \gor \type &
    \text{Kind}\\
    A &\gequal \prd x A A \gor P &
    \text{Type family} \\
    P &\gequal \app {\cst a} S &
    \text{Atomic type} \\
    M &\gequal \lam x M \gor F &
    \text{Canonical object} \\
    F &\gequal \app H S \gor \smeta X \sigma &
    \text{Atomic object} \\
    H &\gequal \var x \gor \cst c \gor \fct{f} &
    \text{Head}\\
    S &\gequal \spinenil \gor \spinecons M S &
    \text{Spine}\\
    \sigma &\gequal \msubstnil \gor \msubstcons \sigma x M &
    \text{Parallel substitution} \\
    \Gamma &\gequal \envnil \gor \envcons \Gamma x A &
    \text{Environment}
  \end{align*}
  \caption{Syntax of \SLF}
  \label{fig:syntax-LF}
\end{figure}

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}. We differ the discussion of
function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
object containing no function symbols. The definition of hereditary
substitution $\gsubst M \sigma$ is skipped and can be found in
\cite{pfenning2007term} (and extended to metavariables as
\cite{nanevski2008contextual}).

@@

@@

The algorithm in \ref{sec:eval} will maintain the invariant that a repository $\mr$ is
\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
$\FMV{\meta X}$ does not contain $\meta X$. If $\mr = (\Delta,
X[\sigma])$, we write $\hat {\mr}$ for the tip $X[\sigma]$, abusively
$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

@@

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
others, and deconstruct them by case analysis on an applicative
term. The definition $T$ of a function always begins with a series of
$\lambda$.

\begin{figure}
  \centering
  \begin{align*}
    T &\gequal \lam x T \gor
    U & \text{Term} \\
    U &\gequal F \gor
    \matchin U \Gamma C & \text{Atomic term}\\
    C &\gequal \casebnobar P U \gor
    C\ \caseb P U & \text{Branches}\\
    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
  \end{align*}

  \caption{Syntax of \CL}
\label{fig:syntax-CL}
\end{figure}

@@

Again, we use indifferently set, map or list notations for signatures.

\begin{definition}
  To each function $\fct f : A = T$ in $\Sigma$ we associate a family
  of \emph{inverses} $\fcti f n : (A)^n = \pi_n(A)$ for $n\in\mathbb
  N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
  \begin{align*}
    (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
    (\app {\cst a} S)^A &= \prd x {\app{\cst a} S} A \\
    \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
    \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
    \pi^y(\app{\cst a} S) &= \var y
  \end{align*}
\end{definition}

@@

\begin{itemize}
\item it evaluates $M$ by reducing function symbols,
\item it type-checks all values fed and produced by functions \wrt\ $\mr$,
\item it slices those values
\end{itemize}

This requires threading the map of $\mr$ throughout evaluation and
typing. We first present the bidirectional typing and slicing part of
the algorithm (Fig. \ref{fig:obj-typing}) as a syntax-directed
inference system. All judgments are of the form $\gjo U V W$ where $U$
and $V$ are inputs, $V$ is the induction hypothesis and $W$ the
output. They are parameterized by a constant and implicit signature
$\Sigma$, and all premises are ordered in the algorithmic order of
computation.

\begin{figure*}

  \fbox{$\jm\Delta\Gamma M A {\Delta'} {M'}$}
  \qquad
  Canonical object

>>>>

\documentclass{llncs}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{ntheorem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{macros}
\usepackage{mathpartir}
\usepackage{url}

\allowdisplaybreaks[1]
\renewcommand{\floatpagefraction}{0.8}

@@

  A lightweight way to design a trusted type-checker is to let it
  return a \emph{certificate} of well-typing, and check it a
  posteriori (for instance \sysname{Agda} and \sysname{GHC} adopt this
  architecture internally).  Most of the time, these type-checkers are
  confronted sequentially with very similar pieces of input: the same
  program, each time slightly modified. Yet they rebuild the entire
  certificates, even for unchanged parts of the code. Instead of
  discarding them between runs, we can take advantage of these
  certificates to improve the reactivity of type-checking, by feeding
  parts of the certificates for unchanged parts of the original
  program back into the new input, thus building an incremental
  type-checking.

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for mere ``type checking'', or explicit
``church-style'' calculi. Take for example dependent type systems,
relying definitional equality on types, or type systems with subtype
polymorphism: both rely on a rule of the form
$$
\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer has to turn the declarative system into an algorithm, and
convince himself that both are equivalent. This task is highly
non-trivial and non-modular: for instance, how to infer the type of a
System \sysname{T} recursor $\recb M N x y P$, in presence of
subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
a typing derivation out of program $M$; it doesn't have to be
complete, or even sound, and \emph{(ii)} a trusted and fast kernel
$\fcheck{\md}$ to verify these derivations \wrt\ a given set of
inference rules $\Sigma$. This is a \emph{certifying} scheme (see
\cite{leroy2006compcert} for a discussion on approaches to reliable
compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$, but the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or return an
incorrect derivation. Explicitly generating the derivation in the
declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

Several generic methods have been devised for turning a batch program
into an incremental one
\cite{pugh1989incremental,acar2003selective}. Unfortunately, they
suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1 = \recb{\s\z}{\s\z} x y {\s {\var x}}$: He sends the command
$\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
well-formed program and tries to construct a valid typing derivation
for its argument in the current, empty context. The system succeeds
and returns a typing derivation $\md_1$ of \/ $\vdash P_1 :
\cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates this derivation, it inserts a coercion $\cst{odd}
\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here. $\md_2$ was deriving $\vdash\s\z:\cst{nat}$ (with a
coercion since it was used as an argument to $\var{add}$); the system
thus infer the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.

% Finally, he decides (in an odd move) to inline the definition of
% $\var{mul}$ into its call site. Call $\md_8$ the derivation of the
% definition of $\var{mul}$ (it depends on an $\var H$ of
% $\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
% $(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
% $\times$. He orders:
% \begin{align*}
% \finfer{}(
% \letb{add}{\fget{\md_5}}
% \fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
%       {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
% )
%   \ \text{.}
% \end{align*}
% The definition for $\var{add}$ is unchanged, but the instantiation of
% open derivations enables to directly substitute one into the other and
% get the appropriate program: call $\md_{10}$ the whole derivation just
% returned, $\fget{\md_{10}}$ returns:
% \begin{align*}
%   &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
%     {\var y} x y {\s{\var x}}}
%   \\&\quad
%   {(\tlam x {\cst{nat}} \tlam y
%     {\cst{nat}} \recb \z {\var y} z t {
%       \app{\var{add}}\app{\var x} {\var z}})}
%   \app{(\app {\var{add}} \app {(\s{\s\z})}
%     {(\s{\s\z})})}{(\s{\s{\s\z}})}
%   \ \text{.}
% \end{align*}

@@

% explicit LF
% lambda-lifting

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solution to this question. It
is a dependently typed lambda-calculus for \emph{representing} terms,
derivations and statements of $\Pi_0^1$ meta-theorems in a given
inference system supporting hypothetical judgments
\cite{pfenning2001logical}. For these, it provides built-in support
for encoding languages with binders and hypothesis (a feature called
$\lambda$-tree syntax or HOAS \cite{pfenning1988higher}). The user
gives a \emph{signature} defining abstract syntax and inference rules
of his \emph{object language}, and gets an algorithm to verify terms
and derivations expressed in the \LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}
This framework has been used to issue certificates for a number of
practical applications like proof-carrying
code~\cite{necula1997proof}, authentication
framework~\cite{appel1999proof} \ldots %TODO ATP

@@

\begin{figure}[t]
  \begin{align*}
    K &\gequal \prd x A K \gor \type &
    \text{Kind}\\
    A &\gequal \prd x A A \gor P &
    \text{Type family} \\
    P &\gequal \app {\cst a} S &
    \text{Atomic type} \\
    M &\gequal \lam x M \gor F &
    \text{Canonical object} \\
    F &\gequal \app H S \gor \smeta X \sigma &
    \text{Atomic object} \\
    H &\gequal \var x \gor \cst c \gor \fct{f} &
    \text{Head}\\
    S &\gequal \spinenil \gor \spinecons M S &
    \text{Spine}\\
    \sigma &\gequal \msubstnil \gor \msubstcons \sigma x M &
    \text{Parallel substitution} \\
    \Gamma &\gequal \envnil \gor \envcons \Gamma x A &
    \text{Environment}
  \end{align*}
  \caption{Syntax of \SLF}
  \label{fig:syntax-LF}
\end{figure}

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995lambda}. We differ the discussion of
function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
object containing no function symbols. The definition of hereditary
substitution $\gsubst M \sigma$ is skipped and can be found in
\cite{pfenning2007term} (and extended to metavariables as
\cite{nanevski2008contextual}).

@@

% TODO relate to LF

@@

If $\mr = (\Delta, X[\sigma])$, we write $\hat {\mr}$ for the tip
$X[\sigma]$, abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$
for $\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
The algorithm in \ref{sec:eval} will maintain the invariant that a
repository $\mr$ is \emph{acyclic}, that is for all $\meta X$ in
$\mr$, the fixpoint of $\FMV{\mr(\meta X)}$ does not contain $\meta X$.

@@

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
others, and deconstruct them by case analysis. The definition $T$ of a function always begins with a series of
$\lambda$.

\begin{figure}[t]
  \begin{align*}
    T &\gequal \lam x T \gor
    U & \text{Term} \\
    U &\gequal F \gor
    \matchin U \Gamma C & \text{Atomic term}\\
    C &\gequal \casebnobar P U \gor
    C\ \caseb P U & \text{Branches}\\
    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern}
  \end{align*}
  \caption{Syntax of \CL}
\label{fig:syntax-CL}
\end{figure}

@@

\begin{definition}
  To each function $\fct f : A = T$ in $\Sigma$ we associate a family
  of \emph{inverses} $\fcti f {\;n} : (A)^n = \pi_n(A)$ for $n\in\mathbb
  N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
  \begin{align*}
    (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
    (\app {\cst a} S)^A &= \prd x {\app{\cst a} S} A \\
    \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
    \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
    \pi^y(\app{\cst a} S) &= \var y
  \end{align*}
\end{definition}

@@

\begin{itemize}
\item it evaluates $M$ by reducing function symbols,
\item it type-checks all values fed to and produced by functions \wrt\
  $\mr$,
\item it slices those values. This requires threading the map of $\mr$
  throughout evaluation and typing.
\end{itemize}

\paragraph{Typing}

We first present the bidirectional typing and slicing part of the
algorithm (Fig. \ref{fig:obj-typing}) as a syntax-directed inference
system. All judgments are of the form $\gjo U V W$ where $U$ and $V$
are inputs, $V$ is the induction hypothesis and $W$ the output. They
are parameterized by a constant and implicit signature $\Sigma$, and
all premises are ordered in the algorithmic order of computation. The
main judgement is atomic object typing $\jf\Delta\Gamma F {\Delta'}
{F'} P$, which reads: in repository $\mr$ and environment $\Gamma$,
applicative term $F$ is well-typed of type $A$, and the slicing of its
value produces a new repository $\mmr$. $F'$ is the \emph{residual} of
this operation: only constant applications $\app{\cst c} S$ are
sliced, $F'$ is the ``tip'' of $F$ that was not sliced.
